---
title: "Obesity-vs.-fast-food-restaurants-by-state"
output: html_document
date: "2023-12-05"
---

```{r}
library(tidyverse)
fastFoodRestaurants <- read_delim("../../fastfood.csv")
spec(fastFoodRestaurants)

obesityByState <- read_delim("../../obesity_by_state.csv")
obesityByState$state = obesityByState$State # prepping data for merge
spec(obesityByState)
```

## Data cleaning & pre-processing

Adding a new categorical variable, whether restaurant is in North or South of the US: 

```{r}
fastFoodRestaurants <- fastFoodRestaurants %>% 
  mutate(
    north = ifelse(latitude >= 37.5 & latitude <= 50, TRUE, FALSE),
    south = ifelse(latitude >= 25 & latitude < 37.65, TRUE, FALSE)
  )
fastFoodRestaurants
```

We already added a new continuous/numerical variable by modifying the dataset itself (calculated the average weight by state (in lbs))


One summarization of data frame, amount of fast food restaurants by city: 

```{r}
summary <- fastFoodRestaurants %>%
  group_by(state) %>%
  summarise(count = n())
summary
```

## Data joining

We are now joining the fastFoodRestaurants dataset and the obesityByState dataset by the state variable. This will enable us to analyze whether obesity rates and the number of fast food restaurants have a correlation and analyze this correlation.

```{r}
joint <- merge(x = fastFoodRestaurants, y = obesityByState, by.x='state', by.y = 'state', fill = -9999) # merging the two big datasets
head(joint, 10)
colnames(joint)
```

```{r}
merged_df <- merge(summary, obesityByState, by = "state")
merged_df
```

##### number of fast food restaurants vs obesity rate
```{r}
library(ggplot2)
ggplot(data = merged_df) +
  geom_point(aes(x = `Obesity `, y = count)) +
  labs(x = "Obesity", y = "Count") +
  ggtitle("Scatterplot of Count vs Obesity") +
  theme_minimal()
```

##### Does the diabetes percentage per state correlate with the obesity rate per state?
```{r}
library(ggplot2)
ggplot(merged_df, aes(x = `Diabetes Percentage`, y = `Obesity `)) +
  geom_point() +
  geom_smooth() +
  labs(title = "Diabetes Percentage vs. Obesity Rate by State",
       x = "Diabetes Percentage",
       y = "Obesity Rate") +
  theme_minimal()
```

#### Is there a correlation between state level rates of obesity and latitude?

#### Blue points are considered North USA, Red points are considered South USA
#### The y axis represents obesity rates per 100,000 people, or the prevalence of obesity in each state
#### Each point on the graph represents a state
```{r}
colors <- ifelse(joint$latitude <= 37.5, "red", "blue")

plot(joint$latitude, joint$`Obesity per 100k`, 
     xlab = "Latitude",
     ylab = "Obesity per 100k",
     main = "Correlation Between Obesity and Latitude",
     col = colors)
```
#### Data correlation between Average Weight vs Obesity rate per state
```{r}
library(ggplot2)

ggplot(obesityByState, aes(x = `Average Weight`, y = `Obesity`)) +
  geom_point() +
  geom_smooth() +
  labs(title = "Average Weight vs. Obesity Rate by State",
       x = "Average Weight",
       y = "Obesity Rate") +
  theme_minimal()

```
As we can see above, the data points are scattered but overall they seem to have a correlation. As every state's average weight increases, so does the obesity rate of the state. We know this because of the linear upturned line demonstrating a upward trend between the two variables. 
